---
title: "Decomposing Headline and Core CPI"
description: "Decompositions of the headline and core CPI indexes"
author: "Owen"
date: "8/6/2025"
categories:
  - CPI
format: 
  html:
    link-external-newwindow: true
---

## Introduction

The first purpose of the decomposition is develop an understand of what is driving changes in headline and core inflation. The second purpose is to have a handy reference.

```{r}
#| label: load packages and data
#| warning: False
#| message: False
#| echo: False

library(readxl)
library(gt)

path <- "C:/Users/Owen/OneDrive/Documents/quarto_test/blog/post-20250806/"

head0 <- readxl::read_excel(paste0(path,"blog_data_20250806.xlsx"),sheet = "Headline_0")

head1 <- readxl::read_excel(paste0(path,"blog_data_20250806.xlsx"),sheet = "Headline_1")

core0 <- readxl::read_excel(paste0(path,"blog_data_20250806.xlsx"),sheet = "Core_0")

housexcle <- readxl::read_excel(paste0(path,"blog_data_20250806.xlsx"),sheet = "Housing_excl_energy")

tranexcle <- readxl::read_excel(paste0(path,"blog_data_20250806.xlsx"),sheet = "Transp_excl_energy")

energy0 <- readxl::read_excel(paste0(path,"blog_data_20250806.xlsx"),sheet = "Energy_0")

energy1 <- readxl::read_excel(paste0(path,"blog_data_20250806.xlsx"),sheet = "Energy_1")

hhenergy <- readxl::read_excel(paste0(path,"blog_data_20250806.xlsx"),sheet = "HH_energy")

energy_comm <- readxl::read_excel(paste0(path,"blog_data_20250806.xlsx"),sheet = "Energy_comm")

food <- readxl::read_excel(paste0(path,"blog_data_20250806.xlsx"),sheet = "Food")


```

Core Inflation excludes energy and food. Food is defined as all food and non-beverages, i.e., it excludes alchoholic beverages.

```{r}
#| label: core and headline
#| warning: False
#| message: False
#| echo: False

head0 %>% gt()

```
We can decompose core CPI, All items less food and energy, as follows:

```{r}
#| echo: False
core0 %>% gt()
```

Housing excluding household energy and transportation excluding energy are derived as follows:

```{r}
#| echo: False

housexcle %>% gt()

tranexcle %>% gt()

```

## Food and Energy in headline inflation

Food is decomposed into food at home and food away from home.
```{r}
#| echo: False

food %>% gt()
```

Energy can be decomposed into energy from commodities and services

```{r}
#| echo: False

energy0 %>% gt()
```

and also as the sum of household energy and motor fuel.

```{r}
#| echo: False

energy1 %>% gt()
```
Household energy is made up of two components
```{r}
#| echo: False

hhenergy %>% gt()
```
There is no further breakdown of energy services that is found within the housing category.

Energy commodities are made up of types of fuel
```{r}
#| echo: False

energy_comm %>% gt()
```





